Network Structure, Efficiency, and Performance in WikiProjects

### 2018-01-12 paper-content.tex
Layout.

\begin{abstract}
The internet has enabled collaborations at a scale never before possible,
but the best practices for organizing such large collaborations are still not clear.
Wikipedia is a visible and successful example of such a collaboration which might offer
insight into what makes large-scale, decentralized collaborations successful.
We analyze the relationship between the structural properties of WikiProject coeditor networks
and the performance and efficiency of those networks.
We also perform numerical simulations to determine whether social learning on an NK model can
reproduce the behavior observed on Wikipedia.
We find several results seen in numerical and small-scale lab studies:
a performance/efficiency trade-off,
higher performance with less skewed node distributions,
and higher performance with shorter path lengths.
We also see behaviors not previously identified: an association between low degree coeditor networks
and both higher performance and higher efficiency,
suggesting possible benefits to decentralized collaborations made of smaller, more tightly-knit teams.
We also propose a novel consensus-based social learning strategy that is both more efficient and higher
performance than existing strategies, and reproduces some behaviors seen in WikiProjects.
\end{abstract}

@@

Our main findings are:
\begin{itemize}
\item WikiProjects having low-degree coeditor networks tend to have both higher performance and higher efficiency, independent of path lengths;
\item Short path lengths tend to be associated with higher performance, consistent with a conformity-based learning strategy;
\item Structural inequality, as measured by degree skewness, is associated with lower performance,
and lower efficiency at reaching the highest quality levels;
\item Using numerical simulations, we show that the performance and efficiency of some networked learning strategies depends on degree distribution.
\end{itemize}

The rest of the paper is structured as follows.
Section \ref{sec:background} reviews related work and the motivation for the present project,
Section \ref{sec:wp} describes the methodology and results of our empirical analysis of WikiProjects,
Section \ref{sec:sim} describes the methodology and results of our numerical simulations,
Section \ref{sec:discuss} discusses the interpretation and broader impact of our results,
and Section \ref{sec:conclusion} concludes.

@@

The field of social learning studies how
groups of interacting agents
can collaborate to solve complex problems.
In {\em networked social learning},
agents are represented by nodes on a network
and can only interact with their neighbors.
Social learning tasks can be divided into cases where agents have {\em generated signals}
(independently noisy estimates of a true value)
and those where agents have {\em interpreted signals}
(solutions based on different selections of available data)
\cite{hong_interpreted_2009}.
The behavior of individual agents is described by their
{\em social learning strategy}.
For generated signals,
a naive Bayesian approach converges to the truth
when all agents have the same degree,
while the speed of convergence depends on the {\em spectral gap}
between the two largest eigenvalues of the network's adjacency matrix
\cite{degroot_reaching_1974,golub_naive_2010}.
Complex social learning tasks can also be modeled as the problem
of maximizing an objective function with many local maxima,
referred to as a {\em rugged landscape}
\cite{lazer_network_2007,mason_propagation_2008,mason_collaborative_2012,grim_scientific_2013,barkoczi_social_2016}.
Numerical simulations have shown that efficient networks
(those with short paths between nodes)
can result in faster convergence at the cost of a less optimal solution,
due to less time for exploration
\cite{mason_propagation_2008,grim_scientific_2013}.
However, when conformity-based social learning strategies are used,
efficient networks can sometimes find more optimal solutions than
inefficient ones \cite{barkoczi_social_2016}.
The above tradeoff between convergence time, optimality, and learning strategy
motivates the distinction between performance and efficiency in the current study.

@@

Research on digital communities has also examined the role of factors
such as diversity and inequality
in collaborative work and decision-making.
Using a simulation model, Hong and Page \cite{hong_groups_2004} found that
diverse groups can outperform groups composed of the best individual
problem-solvers.
In sociology, much research has been done examining the relationship between
network structure and social capital.
Powerful individuals are often ``brokers''
who act as exclusive intermediaries between disconnected portions of the
social network \mbox{\cite{silverman_patronage_1965}}.
Similarly, successful innovation in organizations often occurs in ``structural
holes'' between groups \mbox{\cite{granovetter_strength_1973}}.
There is some evidence that groups with high structural inequality,
as evidenced by highly-skewed degree distributions,
perform worse on collaborative tasks \mbox{\cite{kearns_experiments_2012}}.
In the present project, we use co-editor network degree skewness to
examine the relationships between structural inequality,
project performance, and project efficiency.

@@

We also find that mean min-cut is highly correlated with degree ($r=0.980$, $p<0.001$),
so we exclude min-cut from regression models to prevent collinearity (see Figure \ref{fig:degree-mincut}).
The high correlation between mean degree and min-cut implies that in most cases
the minimum $st$-cut is simply the either set of edges from $s$ or the set of edges to $t$.
The rarity of non-trivial min-cuts suggests that WikiProject coeditor networks have very few central
bottlenecks and are thus highly decentralized.

\begin{figure}
\centering
\includegraphics[width=3in,height=3in]{fig-eff-perf-hist.pdf}
\caption{
Histograms of WikiProject efficiency and performance.
Both measures are highly right-skewed, but form unimodal distributions
with low skewness after log transformation.
\label{fig:eff-perf-hist}
}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=2.33in,height=2.33in]{fig-degree-mincut.pdf}
\caption{
Min-cut of WikiProject coeditor networks vs. their mean degree.
\label{fig:degree-mincut}
}
\end{figure}

@@

@@

\begin{table*}
\small
\centering
\caption{
Social learning strategies
\label{tab:strat}
}
\bigskip
\begin{tabular}{lccccc}
Name          & Social stage & Individual stage & Limited concern & Unknown objective & Single truth \\
\hline
Best+I & Best neighbor   & Global & & & \\
Conf+I & Conformity      & Global & & \Checkmark & \\
Best+LI & Best neighbor  & Local  & \Checkmark & & \\
Conf+LI & Conformity     & Local  & \Checkmark & \Checkmark & \\
LMaj+LI & Local majority & Local  & \Checkmark & \Checkmark & \Checkmark \\
\hline
\end{tabular}
\end{table*}

@@

\begin{table}
\small
\caption{
Degree regression coefficients for agent-based models.
\label{tab:sim-perf}
}
\centering
\bigskip
\begin{tabular}{lll}
Strategy & Perf. Std. Coeff. & Eff. Std. Coeff.\\
\hline
Best+I  & -4.2$\times{10^{-5}}$ & \+4.1$\times{10^{-5}}$ \\
Conf+I  & \+2.7$\times{10^{-5}}$ & \+9.4$\times{10^{-5}}$ \\
Best+LI & -9.6$\times{10^{-4\;**}}$ & \+7.7$\times{10^{-5}}$ \\
Conf+LI & -1.5$\times{10^{-3\;***}}$ & \+8.7$\times{10^{-5\;*}}$ \\
LMaj+LI & \+1.2$\times{10^{-4\;**}}$ & -0.038$^{***}$ \\
\hline
\end{tabular}
\begin{tablenotes}
\item \centering * $p < 0.05$. ** $p < 0.01$. *** $p < 0.001$.
\end{tablenotes}
\end{table}

@@

Figure \ref{fig:val-iter} shows how agents' solutions improve after repeated applications of
different learning strategies and rewiring values.
Each curve represents an average over 100 trials, each with 250 agent solutions.
For all rewiring values, local strategies are less efficient and more performant than their
non-local counterparts.
The consensus strategy is both more efficient and more performant than others,
with its performance increasing with higher rewiring.

The effect of degree on performance and efficiency is shown in Figure \ref{fig:deg-eff-perf} and
Table \ref{tab:sim-perf}.
For both local and non-local versions of the best-neighbor and conformity strategies,
there is no significant effect of degree on performance or efficiency.
As noted above however, the local versions consistently have higher performance and lower efficiency
than their non-local counterparts.
We observe more interesting behavior with the consensus strategy.
Node degree shows a significant negative correlation with degree,
and a small but significant positive correlation with efficiency.
The consensus simulations replicate the efficiency behavior observed in WikiProjects,
but the opposite behavior for performance.
But unlike other strategies, consensus demonstrates that degree distribution can have an
impact on both performance and efficiency.

@@

While existing research into the role of network structure in collaboration
has focused on numerical simulations and lab experiments,
analysis of large real-world systems is an important next step.
Our empirical analysis of WikiProjects contributes several findings towards a better
understanding of large, decentralized, real-world collaboration.
We observe several results consistent with previous work:
a trade-off between performance and efficiency,
higher performance for shorter path lengths in a conformity setting,
and a reduction in performance with increased structural inequality.
By using real-world networks, we were also able to analyze network properties independently,
allowing us to show that while most existing work has focused on the importance of path length,
degree distribution may be just as, or more, important.
The association of low-degree with both high performance and high efficiency is compelling,
as it sidesteps the usual trade-off between performance and efficiency.
In low-degree networks, agents have more repeated interactions with smaller groups of collaborators,
possibly suggesting that small team sizes could be beneficial for large collaborations.
Similarly, the observation that performance is higher in projects with less structural inequality suggests that,
if the challenges of egalitarian organizing can be overcome,
decentralized collaborations may produce better outcomes than those with centralized, top-down structures.

Our numerical simulations offer a few additional insights.
Our proposed local variants of best-neighbor and conformity strategies out-performed
the standard variants while searching fewer alternatives.
The higher performance for less work is analogous to the performance increases that
have been seen for sampling fewer neighbors \cite{barkoczi_social_2016}.
However, our network is a special case in that its structure is related to the structure of
the NK model being optimized, which could increase the usefulness of local information.
Our proposed consensus strategy, intended to model Wikipedia more closely,
exhibits a degree-dependent performance and efficiency, unlike other strategies.
This dependence is also seen in our Wikipedia analysis, although performance's dependence has the opposite sign.
The consensus strategy is also interesting in that it is more efficient and higher performance than the other strategies.
The unique properties of the consensus strategy may be due to a major difference it has from the others:
rather than exploring through local hill-climbing,
the consensus strategy combines parts of existing solutions to create entirely new ones.

Our work has several limitations.
Our empirical analysis is purely correlative and cannot be used to draw
conclusions about the causal influence of network structure on collaboration.
However, the consistency of our results with other lab-based and numerical studies
suggests that the causal link is worth further study.
Similarly, our study focuses entirely on a single setting and while it is suggestive,
does not necessarily generalize,
and studies of other systems are necessary to corroborate our findings.

@@

In this paper, we have described the relationship between the structural properties of WikiProject
coeditor networks and the performance and efficiency of those networks.
As in other studies, we see a trade-off between performance and efficiency,
but not an absolute trade-off.
Some properties, such as low degree, are associated with both higher performance and higher efficiency.
We also find that the correlations between path length and performance are consistent with a conformity-based
social learning strategy, but not a greedy best-neighbor strategy.
We also observe improved performance in more decentralized project, as has been seen in small-scale lab experiments.
While the performance and efficiency of most previously-studied social learning strategies depend more on path length than
degree,
we have proposed a novel consensus learning strategy that is more realistic, more efficient, and higher performance than
existing strategies.
While the efficiency and performance of existing strategies do not exhibit any degree-dependence, our consensus strategy
demonstrates that degree-dependence is possible.
While additional work is needed to determine causal relationships and how generalizable our results are,
we have shown evidence that several phenomena predicted by numerical and small-scale lab experiments are present in a large,
real-world network.
And, we have identified new behaviors not explained by existing models, suggesting the need for alternate models.
Our results suggest that the success of large-scale collaborations may be aided by greater decentralization,
consensus or conformity-based decision-making, and more tightly-knit collaborations between smaller teams.

>>>>

\begin{abstract}
The internet has enabled collaborations at a scale never before possible,
but the best practices for organizing such large collaborations are still not clear.
Wikipedia is a visible and successful example of such a collaboration which might offer
insight into what makes large-scale, decentralized collaborations successful.
We analyze the relationship between the structural properties of WikiProject coeditor networks
and the performance and efficiency of those networks.
We also perform agent-based simulations to determine whether social learning on an NK model can
reproduce the behavior observed on Wikipedia.
We find several results seen in numerical and small-scale lab studies:
a performance/efficiency trade-off,
higher performance with less skewed node distributions,
and higher performance with shorter path lengths.
We also see behaviors not previously identified: an association between low degree coeditor networks
and both higher performance and higher efficiency,
suggesting possible benefits to decentralized collaborations made of smaller, more tightly-knit teams.
We also propose a novel local majority-based social learning strategy that is both more efficient and higher
performance than existing strategies, and reproduces some behaviors seen in WikiProjects.
\end{abstract}

@@

Our main findings are:
\begin{itemize}
\item WikiProjects having low-degree coeditor networks tend to have both higher performance and higher efficiency, independent of path lengths;
\item Short path lengths tend to be associated with higher performance, consistent with a conformity-based learning strategy;
\item Structural inequality, as measured by degree skewness, is associated with lower performance,
and lower efficiency at reaching the highest quality levels;
\item Using agent-based modeling, we show that the performance and efficiency of some networked learning strategies depends on degree distribution.
\end{itemize}

The rest of the paper is structured as follows.
Section \ref{sec:background} reviews related work and the motivation for the present project,
Section \ref{sec:wp} describes the methodology and results of our empirical analysis of WikiProjects,
Section \ref{sec:sim} describes the methodology and results of our agent-based models,
Section \ref{sec:discuss} discusses the interpretation and broader impact of our results,
and Section \ref{sec:conclusion} concludes.

@@

The field of social learning studies how
groups of interacting agents
can collaborate to solve complex problems.
In {\em networked social learning},
agents are represented by nodes on a network
and can only interact with their neighbors.
Social learning tasks can be divided into cases where agents have {\em generated signals}
(independently noisy estimates of a true value)
and those where agents have {\em interpreted signals}
(solutions based on different selections of available data)
\cite{hong_interpreted_2009}.
The behavior of individual agents is described by their
{\em social learning strategy}.
For generated signals,
a naive Bayesian approach converges to the truth
when all agents have the same degree,
while the speed of convergence depends on the {\em spectral gap}
between the two largest eigenvalues of the network's adjacency matrix
\cite{degroot_reaching_1974,golub_naive_2010}.
Complex social learning tasks can also be modeled as the problem
of maximizing an objective function with many local maxima,
referred to as a {\em rugged landscape}
\cite{lazer_network_2007,mason_propagation_2008,mason_collaborative_2012,grim_scientific_2013,barkoczi_social_2016}.
Numerical simulations have shown that efficient networks
(those with short paths between nodes)
can result in faster convergence at the cost of a less optimal solution,
due to less time for exploration
\cite{mason_propagation_2008,grim_scientific_2013}.
However, when conformity-based social learning strategies are used,
efficient networks can sometimes find more optimal solutions than
inefficient ones \cite{barkoczi_social_2016}.
The above trade-off between convergence time, optimality, and learning strategy
motivates the distinction between performance and efficiency in the current study.

@@

Research on digital communities has also examined the role of factors
such as diversity and inequality
in collaborative work and decision-making.
Using a simulation model, Hong and Page \cite{hong_groups_2004} found that
diverse groups can outperform groups composed of the best individual
problem-solvers.
In sociology, much research has been done examining the relationship between
network structure and social capital.
Powerful individuals are often ``brokers''
who act as exclusive intermediaries between disconnected portions of the
social network \mbox{\cite{silverman_patronage_1965}}.
Similarly, successful innovation in organizations often occurs in ``structural
holes'' between groups \mbox{\cite{granovetter_strength_1973}}.
There is some evidence that groups with high structural inequality,
as evidenced by highly-skewed degree distributions,
perform worse on collaborative tasks \cite{kearns_experiments_2012}.
In the present project, we use co-editor network degree skewness to
examine the relationships between structural inequality,
project performance, and project efficiency.

@@

We also find that mean min-cut is highly correlated with degree ($r=0.980$, $p<0.001$),
so we exclude min-cut from regression models to prevent collinearity.
The high correlation between mean degree and min-cut implies that in most cases
the minimum $st$-cut is simply the either set of edges from $s$ or the set of edges to $t$.
The rarity of non-trivial min-cuts suggests that WikiProject coeditor networks have very few central
bottlenecks and are thus highly decentralized.

\begin{figure}
\centering
\includegraphics[width=2.5in,height=2.5in]{fig-eff-perf-hist.pdf}
\caption{
Histograms of WikiProject efficiency and performance.
Both measures are highly right-skewed, but form unimodal distributions
with low skewness after log transformation.
\label{fig:eff-perf-hist}
}
\end{figure}

%\begin{figure}
%\centering
%\includegraphics[width=2.33in,height=2.33in]{fig-degree-mincut.pdf}
%\caption{
%Min-cut of WikiProject coeditor networks vs. their mean degree.
%\label{fig:degree-mincut}
%}
%\end{figure}

@@

\begin{table*}[!ht]
\small
\centering
\caption{
Social learning strategies
\label{tab:strat}
}
\bigskip
\begin{tabular}{lccccc}
Name          & Social stage & Individual stage & Limited concern & Unknown objective & Single truth \\
\hline
Best+I & Best neighbor   & Global & & & \\
Conf+I & Conformity      & Global & & \Checkmark & \\
Best+LI & Best neighbor  & Local  & \Checkmark & & \\
Conf+LI & Conformity     & Local  & \Checkmark & \Checkmark & \\
LMaj+LI & Local majority & Local  & \Checkmark & \Checkmark & \Checkmark \\
\hline
\end{tabular}
\end{table*}

@@

@@

\begin{table}
\small
\caption{
Degree regression coefficients for simulations.
\label{tab:sim-perf}
}
\centering
\bigskip
\begin{tabular}{lll}
Strategy & Perf. Std. Coeff. & Eff. Std. Coeff.\\
\hline
Best+I  & -4.2$\times{10^{-5}}$ & \+4.1$\times{10^{-5}}$ \\
Conf+I  & \+2.7$\times{10^{-5}}$ & \+9.4$\times{10^{-5}}$ \\
Best+LI & -9.6$\times{10^{-4\;**}}$ & \+7.7$\times{10^{-5}}$ \\
Conf+LI & -1.5$\times{10^{-3\;***}}$ & \+8.7$\times{10^{-5\;*}}$ \\
LMaj+LI & \+1.2$\times{10^{-4\;**}}$ & -0.038$^{***}$ \\
\hline
\end{tabular}
\begin{tablenotes}
\item \centering * $p < 0.05$. ** $p < 0.01$. *** $p < 0.001$.
\end{tablenotes}
\end{table}

@@

Figure \ref{fig:val-iter} shows how agents' solutions improve after repeated applications of
different learning strategies and rewiring values.
Each curve represents an average over 100 trials, each with 250 agent solutions.
For all rewiring values, local strategies are less efficient and more performant than their
non-local counterparts.
For the best-neighbor strtegy, local outperforms global.
For local conformity strategy performs notably worse than all other strategies.
The local majority strategy is both more efficient and more performant than others,
with its performance increasing with higher rewiring.
These results imply that,
at least in a simple model of collaboration,
the performance and efficiency can be simultaneously increased.
Furthermore, performance and efficiency are potentially affected by both the choice of
learning strategy and the average degree of the agents' social network.

The effect of degree on performance and efficiency is shown in Figure \ref{fig:deg-eff-perf} and
Table \ref{tab:sim-perf}.
For non-local versions of both conformity and best-neighbor strategy,
there is no significant effect of degree on performance or efficiency.
The local best-neighbor strategy shows reduced performance with increasing degree,
but no change in efficiency.
Local conformity and local majority show opposite behavior as degree increases:
with local conformity gaining efficiency at the expense of performance,
while local majority increases in performance and decreases in efficiency.
The largest effect size is achieved for efficiency in the local majority simulation,
which is consistent with the efficiency behavior observed in WikiProjects.
However, the performance behavior for local majority is opposite that observed on
Wikipedia.
These agent-based models confirm that network degree has the potential to influence
the performance and efficiency of collaborations.
Furthermore, this influence can be drastically different depending on the strategies
used by collaborators.

@@

While existing research into the role of network structure in collaboration
has focused on numerical simulations and lab experiments,
analysis of large real-world systems is an important next step.
Our empirical analysis of WikiProjects contributes several findings towards a better
understanding of large, decentralized, real-world collaboration.
We observe several results consistent with previous work:
a trade-off between performance and efficiency
\cite{mason_propagation_2008,grim_scientific_2013},
higher performance for shorter path lengths in a conformity setting
\cite{barkoczi_social_2016},
and a reduction in performance with increased structural inequality
\cite{kearns_experiments_2012}.
By using real-world networks, we were also able to analyze network properties independently,
allowing us to show that while most existing work has focused on the importance of path length,
degree distribution may be just as, or more, important.
The association of low-degree with both high performance and high efficiency is compelling,
as it sidesteps the usual trade-off between performance and efficiency.
In low-degree networks, agents have more repeated interactions with smaller groups of collaborators,
possibly suggesting that small team sizes could be beneficial for large collaborations.
Similarly, the observation that performance is higher in projects with less structural inequality suggests that,
if the challenges of egalitarian organizing can be overcome,
decentralized collaborations may produce better outcomes than those with centralized, top-down structures.

Our agent-based models offer a few additional insights.
We observe degree-dependent performance and efficiency dependence for both local conformity
and local majority strategies.
However, these two strategies have opposite degree dependence,
suggesting that different strategies may be preferable for high-degree and low-degree
networks.
Our local majority strategy,
designed to satisfy several properties present in real-world collaborations,
shows the strongest effects on performance and efficiency as network degree changes.
For the local majority strategy,
the relationship between degree and efficiency is consistent with our
empirical observations on Wikipedia,
suggesting one possible mechanism underlying that efficiency dependence.
However, the performance dependence of this strategy
is opposite that observed on Wikipedia,
suggesting that either the local majority strategy
is incompatible with actual behavior on Wikipedia or
that other factors (possibly including degree skewness)
outweigh the contribution of mean degree.

Our work has several limitations.
Our empirical analysis is purely correlative and cannot be used to draw
conclusions about the causal influence of network structure on collaboration.
However, the consistency of our results with other lab-based and numerical studies
suggests that the causal link is worthy of further study.
Similarly, our study focuses entirely on a single online community and while it is suggestive,
does not necessarily generalize,
and studies of other systems are necessary to corroborate our findings.

@@

In this paper, we have described the relationship between the structural properties of WikiProject
coeditor networks and the performance and efficiency of those networks.
As in other studies, we see a trade-off between performance and efficiency,
but not an absolute trade-off.
Some properties, such as low degree, are associated with both higher performance and higher efficiency.
We also find that the correlations between path length and performance are consistent with a conformity-based
social learning strategy, but not a greedy best-neighbor strategy.
We also observe improved performance in more decentralized project, as has been seen in small-scale lab experiments.
While the performance and efficiency of most previously-studied social learning strategies depend more on path length than
degree,
we have proposed a novel local majority learning strategy that is more realistic, more efficient, and higher performance than
existing strategies.
We observe that in both the local majority strategy and a localized version of
the conformity strategy,
the performance and efficiency of a collaboration can exhibit degree dependence,
and that the direction of that dependence depends on the specific strategy being used.
While additional work is needed to determine causal relationships and how generalizable our results are,
we have shown evidence that several phenomena predicted by numerical and small-scale lab experiments are present in a large,
real-world network.
And, we have identified new behaviors not explained by existing models, suggesting the need for alternate models.
Our results suggest that the success of large-scale collaborations may be aided by greater decentralization,
consensus or conformity-based decision-making, and more tightly-knit collaborations between smaller teams.
